---
title: "Statistical Calibration of Species Co-occurrence in CAPTAIN"
subtitle: "Sub-project 2: HMSC Latent Variable Extraction"
author: "Earth and Environmental Science, Lund University"
date: "2026-06-02"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
---

## Introduction and Conceptual Framework

A central problem in ecological modeling is incorporating species co-occurrence patterns observed in empirical datasets into process-based frameworks like CAPTAIN. Joint Species Distribution Models (JSDMs), specifically Hierarchical Modelling of Species Communities (HMSC), separate these patterns into environmental niches (fixed effects) and residual associations (random effects).

The residual variance-covariance matrix $\Omega$ is modeled via latent variables at the sampling-unit level using site scores ($\eta$) and species scores ($\lambda$). The linear predictor vector for site $i$ and species $j$ is formulated as:

$$L_{ij} = L_{ij}^F + L_{ij}^R$$

where $L_{ij}^F$ represents the fixed environmental layout, and the random component is defined as:

$$L_{ij}^R = \sum_{h=1}^{n_f} \eta_{ih} \lambda_{hj}$$

This architecture means that $\Omega = \Lambda^T \Lambda$ serves as a data-driven summary of co-variation, which may capture unmeasured environmental gradients, micro-refugia, or authentic biological interactions such as competitive exclusion.

---

### Step 1: Simulating Community Matrix with Hidden Structuring

We simulate a community composed of two species across 200 spatial patches. Both species respond identically to a single known environmental driver ($x_1$). However, we introduce an unmeasured environmental gradient ($\eta$) to which the species respond with opposite signs, generating an apparent pattern of competitive exclusion in the residuals.

```{r}
#| label: simulation-setup
#| message: false
#| warning: false

set.seed(42)
n_sites <- 200

# Continuous environmental covariate (Fixed Effect)
x1 <- rnorm(n_sites)
XData <- data.frame(x1 = x1)
X <- cbind(rep(1, n_sites), x1)

# True parameters for the environmental niches (Beta)
beta <- matrix(c(0.5, 1.2,   # Sp 1
                 0.5, 1.2),  # Sp 2
               nrow = 2, byrow = TRUE)

L_fixed <- X %*% t(beta)

# Generate the hidden environmental gradient (True Latent Factor)
eta_true <- rnorm(n_sites, mean = 0, sd = 1)

# Assign opposing species scores (Lambda) to enforce negative co-occurrence
lambda_true <- matrix(c(0.1,   # Species 1
                        0.7), # Species 2
                      nrow = 2, byrow = TRUE)

L_random <- eta_true %*% t(lambda_true)
L <- L_fixed + L_random

# Apply probit transformation to obtain binary presence-absence matrix (Y)
noise <- matrix(rnorm(n_sites * 2), ncol = 2)
Y <- 1 * ((L + noise) > 0)
colnames(Y) <- c("sp1", "sp2")
```


```{r}
#| label: simulation-diagnostics
colMeans(Y)
table(Y[, 1], Y[, 2])
```
### Step 2: Model Specifications and MCMC Fitting via HMSC

We fit an unstructured sampling-unit level random effect using the Hmsc package. The model is deliberately blind to the second gradient, forcing the variance into the latent parameter layers.

```{r}
#| label: model-fitting
#| message: false
#| warning: false
#| results: 'hide'

library(Hmsc)

studyDesign <- data.frame(sample = as.factor(1:n_sites))
rL <- HmscRandomLevel(units = studyDesign$sample)
XFormula <- ~ x1

model <- Hmsc(Y = Y, XData = XData, XFormula = XFormula,
              studyDesign = studyDesign, ranLevels = list(sample = rL),
              distr = "probit")

# Markov Chain Monte Carlo parameter definitions
thin <- 10
samples <- 250
transient <- 500
nChains <- 2

m_fit <- sampleMcmc(model, thin = thin, samples = samples, 
                    transient = transient, nChains = nChains, 
                    verbose = FALSE)
```
### Step 3: Extracting Latent Mathematical Layers

To calibrate CAPTAIN, we must systematically extract the mean posterior values for the matrices representing the species responses to unmeasured gradients ($\Lambda$), the hidden site values ($\eta$), and the resulting residual correlation matrix ($\Omega$).

```{r}
#| label: parameter-extraction

# 3.1 Extract the Residual Correlation Matrix (Omega)
omega_list <- computeAssociations(m_fit)
residual_corr <- omega_list[[1]]$mean
print(residual_corr)

# 3.2 Extract Species Scores (Lambda Matrix)
lambda_post <- getPostEstimate(m_fit, "Lambda")
estimated_lambda <- lambda_post$mean
print(estimated_lambda)

# 3.3 Extract Site Scores (Eta Matrix)
eta_post <- getPostEstimate(m_fit, "Eta")
estimated_eta <- eta_post$mean
head(estimated_eta)
```

Visualizing the Residual Correlation: We plot the extracted correlation structure to confirm that the model isolated the hidden pattern of negative association between the two simulated species.

```{r}
#| label: plot-associations
#| fig-cap: "Estimated residual correlation matrix showing strong negative association."
library(corrplot)
corrplot(residual_corr, method = "color", col = c("grey", "white", "black"))
```

### Step 4: Conditional vs Unconditional Forecasting Configurations

This section demonstrates the difference in predictive capability when conditioning the presence probability of a target species on known data from co-occurring species within the community.

```{r}
#| label: forecasting-simulations

# 4.1 Unconditional Predictions (Baseline Environmental Projection)
uncond_list <- predict(m_fit, expected = TRUE)
uncond_preds <- apply(simplify2array(uncond_list), c(1, 2), mean)

# 4.2 Conditional Predictions (Leveraging the Interspecific Association)
#partition <- createPartition(m_fit, nfolds = 2, column = "sample")
partition.sp <- c(1, 2) # Cross-condition species 1 and species 2

cond_list <- computePredictedValues(m_fit, #partition = partition, 
                                    partition.sp = partition.sp, 
                                    mcmcStep = 100)
cond_preds <- apply(simplify2array(cond_list), c(1, 2), mean)
```
Analytical Comparison: We format the outputs into a structured frame to track how knowing the state of Species 2 drastically alters the assigned probability of Species 1 relative to the unconditional baseline model.

```{r}
#| label: output-table
comparison_df <- data.frame(
  Observed_Sp1 = Y[, 1],
  Observed_Sp2 = Y[, 2],
  Unconditional_Prob_Sp1 = round(uncond_preds[, 1], 3),
  Conditional_Prob_Sp1 = round(cond_preds[, 1], 3)
)

head(comparison_df, 15)
```

